---
title: "2013 Influenza A H7N9 cases in China"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning=FALSE, message=FALSE)
```


This dataste is the the 2013 outbreak of influenza A H7N9 in China, which provided by outbreaks package.

The purpose of this project is to predict disease outcome.

```{r}
# load Data and Libraries
library(outbreaks)
dataset <- fluH7N9.china.2013
```

## Data exploration

As we can see the summary below, this dataset contains case ID, date of onset, date of hospitalisation, date of outcome, gender, age, province and outcome attributes. In addiction to that, there are a lot of missing data. I will deal with these values latter.

There are 57 cases with unknown outcome. Therefore, this part will be divided into test data.

```{r}
summary(dataset)
```

```{r}
str(dataset)
```

We can see that age have some ? values. We should convert ? values to NA

```{r}
# convert age ? to NAs
dataset$age[which(dataset$age == "?")] <- NA
```

Before cleaning dataset,let??s look at scatterplots of all pairs of attributes

First, I create datset without NA outcome as test data cannot be used for exploration

```{r}
dataset_gather <- dataset[!is.na(dataset$outcome),]
```

```{r fig.width = 9, fig.height = 9}
library(GGally)
pairs(dataset_gather)
```

#### Gender

The graph below shows the relationship between gender and outcome
```{r}
# add a level for unknown gender
levels(dataset_gather$gender) <- c(levels(dataset_gather$gender), "unknown")
dataset_gather$gender[is.na(dataset_gather$gender)] <- "unknown"

# rename gender
library(plyr)
dataset_gather$gender <- mapvalues(dataset_gather$gender, from = c("m", "f", "unknown"), 
                                   to = c("Male", "Female", "Unknown gender"))

library(ggplot2)
ggplot(data = dataset_gather, aes(x = gender, fill = outcome, color = outcome))+
    geom_bar(position = "dodge", alpha = 0.7, size = 1) +
    scale_fill_brewer(palette="Set1", na.value = "grey50") +
    scale_color_brewer(palette="Set1", na.value = "grey50") +
    labs(
        y = "Count",
        title = "2013 Influenza A H7N9 cases in China",
        subtitle = "Gender and Province numbers of flu cases"
    )

```

It is obvious that there are more male than female cases and more deaths, recoveries and unknown outcomes in men than in women.


#### Age
Next thing it might be interesting to see the age distribution

```{r}
ggplot(data = dataset_gather[!is.na(dataset_gather$age), ], aes(x = as.numeric(age), fill = outcome, color = outcome)) +
  geom_density(alpha = 0.3, size = 1) +
  geom_rug() +
  scale_color_brewer(palette="Set1", na.value = "grey50") +
  scale_fill_brewer(palette="Set1", na.value = "grey50") +
  labs(
    x = "Age",
    y = "Density",
    subtitle = "Age distribution of flu cases"
  )
```

From the density curve above, We can see that people who died tended to be slightly older than those who recovered.


#### Province

Next thing it might be interesting to see if there is any relationship between province and outcome. I only keep Jiangsu, Shanghai and Zhejian and separate provinces and combine other categories to other province as most provinces have only a handful of cases

```{r}
# rename provinces
dataset_gather$province <- mapvalues(dataset_gather$province, 
                            from = c("Anhui", "Beijing", "Fujian", "Guangdong", "Hebei", "Henan", "Hunan", "Jiangxi", "Shandong", "Taiwan"), 
                            to = rep("Other province", 10))

# rearrange province order so that Other is the last
dataset_gather$province <- factor(dataset_gather$province, levels = c("Jiangsu",  "Shanghai", "Zhejiang", "Other province"))

ggplot(data = dataset_gather, aes(x = province, fill = outcome, color = outcome))+
    geom_bar(position = "dodge", alpha = 0.7, size = 1) +
    scale_fill_brewer(palette="Set1", na.value = "grey50") +
    scale_color_brewer(palette="Set1", na.value = "grey50") +
    labs(
        y = "Count",
        title = "2013 Influenza A H7N9 cases in China",
        subtitle = "Gender and Province numbers of flu cases"
    )
```

Most unknown outcomes were recorded in Zhejiang and Shanghai has the hightest death cases.

#### Date
To see how many days passed between onset, hospitalisation and outcome for each case, let's plot the dates form each cases.


```{r}
# gather for plotting with ggplot2
library(tidyr)
dataset_gather_2 <- dataset_gather %>%
    gather(Group, Date, date.of.onset:date.of.outcome)

# rearrange group order
dataset_gather_2$Group <- factor(dataset_gather_2$Group, levels = c("date.of.onset", "date.of.hospitalisation", "date.of.outcome"))

# rename groups
dataset_gather_2$Group <- mapvalues(dataset_gather_2$Group, from = c("date.of.onset", "date.of.hospitalisation", "date.of.outcome"), 
            to = c("Date of onset", "Date of hospitalisation", "Date of outcome"))

ggplot(data = dataset_gather_2, aes(x = Date, y = as.numeric(age), color = outcome)) +
    geom_point(aes(shape = gender), size = 1.5, alpha = 0.6) +
    geom_path(aes(group = case.ID)) +
    facet_wrap( ~ province, ncol = 2) +
    scale_shape_manual(values = c(15, 16, 17)) +
    scale_color_brewer(palette="Set1", na.value = "grey50") +
    scale_fill_brewer(palette="Set1") +
    labs(
        color = "Outcome",
        shape = "Gender",
        x = "Month in 2013",
        y = "Age",
        title = "2013 Influenza A H7N9 cases in China",
        subtitle = "Dataset from 'outbreaks' package (Kucharski et al. 2014)",
        caption = "\nTime from onset of flu to outcome."
    )

```

It is hard to draw a general conclusion as there are many missing values in the dates. Therefore, I will dig into the days between onset and outcome and between onset and hospitalisation.
 
```{r}
#calculate days_onset_to_outcome
dataset_gather$days_onset_to_outcome <- as.numeric(as.character(gsub(" days", "",
                               as.Date(as.character(dataset_gather$date.of.outcome), format = "%Y-%m-%d") - 
                               as.Date(as.character(dataset_gather$date.of.onset), format = "%Y-%m-%d"))))

#calculate days_onset_to_hospital
dataset_gather$days_onset_to_hospital <- as.numeric(as.character(gsub(" days", "",
                        as.Date(as.character(dataset_gather$date.of.hospitalisation), format = "%Y-%m-%d") - 
                        as.Date(as.character(dataset_gather$date.of.onset), format = "%Y-%m-%d"))))

p1 <- ggplot(data = dataset_gather[!is.na(dataset_gather$days_onset_to_outcome), ], aes(x = days_onset_to_outcome, fill = outcome, color = outcome)) +
    geom_density(alpha = 0.3, size = 1) +
    geom_rug() +
    scale_color_brewer(palette="Set1", na.value = "grey50") +
    scale_fill_brewer(palette="Set1", na.value = "grey50") +
    labs(
        x = "Days_onset_to_outcome",
        y = "Density"
    )

p2 <- ggplot(data = dataset_gather[!is.na(dataset_gather$days_onset_to_hospital), ], aes(x = days_onset_to_hospital, fill = outcome, color = outcome)) +
    geom_density(alpha = 0.3, size = 1) +
    geom_rug() +
    scale_color_brewer(palette="Set1", na.value = "grey50") +
    scale_fill_brewer(palette="Set1", na.value = "grey50") +
    labs(
        x = "Days_onset_to_hospital",
        y = "Density"
    )

library(gridExtra)
library(grid)

grid.arrange(p1, p2, ncol = 2)
```

As we can see from the first density curve, people who have this disease will die in a short period of time. If the patients cannot recover within the specified period of time, they are more likely to die. 

The second density curve shows that shortening the days between onset and hospitalisation can decrease the death rate

Next, I want to plot whether early onset patients have different outcome

```{r}
dataset_gather$early_onset = as.factor(ifelse(dataset_gather$date.of.onset < summary(dataset_gather$date.of.onset)[[3]], 'early onset', 'not early onset'))

ggplot(data = dataset_gather[!is.na(dataset_gather$early_onset), ], aes(x = early_onset, fill = outcome, color = outcome))+
    geom_bar(position = "dodge", alpha = 0.7, size = 1) +
    scale_fill_brewer(palette="Set1", na.value = "grey50") +
    scale_color_brewer(palette="Set1", na.value = "grey50") +
    labs(
        x = "Coset type",
        y = "Count",
        title = "2013 Influenza A H7N9 cases in China"
    )
```

As expected, early onset patients have higher number of death.


## Data wrangling

```{r}
#convert gender into numeric values with 1 for female and 0 for male
dataset$gender_f <- as.factor(ifelse(dataset$gender == "f", 1, 0))

#convert provinces to binary classifiers (yes == 1, no == 0) for Shanghai, Zhejiang, Jiangsu and other provinces
dataset$province_Jiangsu = as.factor(ifelse(dataset$province == "Jiangsu", 1, 0))
dataset$province_Shanghai = as.factor(ifelse(dataset$province == "Shanghai", 1, 0))
dataset$province_Zhejiang = as.factor(ifelse(dataset$province == "Zhejiang", 1, 0))
dataset$province_other = as.factor(ifelse(dataset$province == "Zhejiang" | dataset$province == "Jiangsu" | dataset$province == "Shanghai", 0, 1))

#calculate days_onset_to_outcome
dataset$days_onset_to_outcome <- as.numeric(as.character(gsub(" days", "",
                               as.Date(as.character(dataset$date.of.outcome), format = "%Y-%m-%d") - 
                               as.Date(as.character(dataset$date.of.onset), format = "%Y-%m-%d"))))

#calculate days_onset_to_hospital
dataset$days_onset_to_hospital <- as.numeric(as.character(gsub(" days", "",
                        as.Date(as.character(dataset$date.of.hospitalisation), format = "%Y-%m-%d") - 
                        as.Date(as.character(dataset$date.of.onset), format = "%Y-%m-%d"))))


#the same binary classification is given for whether a case was hospitalised, and whether they had an early onset or early outcome 
dataset$hospital = as.factor(ifelse(is.na(dataset$date.of.hospitalisation), 0, 1))

dataset$early_onset = as.factor(ifelse(dataset$date.of.onset < summary(dataset$date.of.onset)[[3]], 1, 0))

dataset$early_outcome = as.factor(ifelse(dataset$date.of.outcome < summary(dataset$date.of.outcome)[[3]], 1, 0))

#change age data type
dataset$age = as.numeric(dataset$age)

dataset <- subset(dataset, select = -c(2:4, 6, 8))
```


### Imputing missing data

Now we need to check for missing values and look the number and proportion of missing values for each variable using the sapply() function.

```{r}
#check all missing values 
sapply(dataset,function(x) sum(is.na(x)))

pMiss <- function(x){sum(is.na(x))/length(x)*100}
apply(dataset,2,pMiss)
```

We can see that date.of.hospitalisation and date.of.outcome are missing more than 45 percentage, therefore we consider to Use mice package for looking at missing data pattern. 

```{r}
library(mice)
md.pattern(dataset[, -c(2)])
```

The output tells us that 42 samples are complete, 27 samples miss only the date.of.hospitalisation attribute, 20 samples miss only the date.of.outcome attribute and so on.

The mice() function takes care of the imputing process

```{r}
dataset_impute <- mice(dataset[, -c(2)], print = FALSE, seed=500)
summary(dataset_impute)
```

Next, we can the complete() function to fill in missing values and merge the completed dataset with outcome column

```{r}
dataset_complete <- complete(dataset_impute,1)
dataset_complete <- merge(dataset[, c(1,2), drop = FALSE], dataset_complete, by = "case.ID", all = TRUE)

#remove case.ID attribute
dataset_complete <- dataset_complete[, -1]

summary(dataset_complete$days_onset_to_outcome)

```

## Build Models

I split the imputed data frame into training and test data. Test data are the 57 cases with unknown outcome.

```{r}
# split the dataset into train and test
train_index <- which(is.na(dataset_complete$outcome))
train_data <- dataset_complete[-train_index, ]
test_data  <- dataset_complete[train_index, -1]

library(caret)
# Run algorithms using 10-fold cross validation
control <- trainControl(method="cv", number=10)
```

The random number seed should be reset before reach run to ensure that the evaluation of each algorithm is performed using exactly the same data splits. It ensures the results are directly comparable

### Classification and Regression Trees

```{r}
set.seed(27)
fit.cart <- train(outcome ~ .,
                  data=train_data, 
                  method="rpart", 
                  trControl=control)

```

The graph below shows which feature contributes to the prediction of the outcome
```{r}
library(rattle)
fancyRpartPlot(fit.cart$finalModel)
```


### Random Forest
```{r}
set.seed(27)
fit.rf <- train(outcome ~ .,
                data=train_data, 
                method="rf", 
                trControl=control)
```

#### Plot importance
```{r}
# estimate variable importance
importance <- varImp(fit.rf, scale=TRUE)

# plot importance
importance_df <- importance$importance
importance_df$group <- rownames(importance_df)

library(plyr)
importance_df$group <- mapvalues(importance_df$group, 
                from = c("age", "hospital2", "gender_f2", "province_Jiangsu2", "province_Shanghai2", "province_Zhejiang2", "province_other2", "days_onset_to_outcome", "days_onset_to_hospital", "early_onset2", "early_outcome2" ),
               to = c("Age", "Hospital", "gender", "Jiangsu", "Shanghai", "Zhejiang", "Other province", "Days onset to outcome", "Days onset to hospital", "Early onset", "Early outcome" ))


ggplot() +
  geom_point(data = importance_df, aes(x=Overall, y=reorder(group, Overall))) +
  labs(
    x = "Importance",
    title = "2013 Influenza A H7N9 cases in China",
    subtitle = "Scaled feature importance",
    caption = "\nDetermined with Random Forest and
    repeated cross validation (10 repeats, 10 times)"
  )
```

This graph shows that age is the most important determining factor for predicting disease outcome, followed by days between onset an outcome, days between onset and hospitalisation, and early outcome.

### Naive Bayes
```{r}
set.seed(27)
fit.nb <- train(outcome ~ .,
                data=train_data, 
                method="nb", 
                trControl=control)

```

### Boosting
```{r}
set.seed(27)
fit.bs <- train(outcome ~ .,
                data=train_data, 
                method="ada", 
                trControl=control)

```

## Select Best Model

We can report on the accuracy of each model by creating a list of the created models and using the summary function.

```{r}
#summarize accuracy of models
results <- resamples(list(cart=fit.cart, rf=fit.rf, rf=fit.rf, nb=fit.nb, bs=fit.bs))
summary(results)
```

Next, I create a plot of the model evaluation results and compare the spread and the mean accuracy of each model.

We can see that the most accurate model in this case was random forest:
```{r}
# compare accuracy of models
dotplot(results)
```

The results for the random forest model can be summarized.
```{r}
# summarize Best Model
fit.rf$results
```

This gives a nice summary of what was used to train the model and the mean and standard deviation (SD) accuracy achieved.


Let's take a look at the confusion matrix.
```{r}
#confusion matrix
fit.rf$finalModel$confusion
```

Finally, we can use this model to predect test data
```{r}
# predict on test data
Prediction <- predict(fit.rf, test_data)

test_data$outcome <- Prediction
```

## Conclusion

This dataset does not have enough data to reliably predict the outcome. Therefore, more cases should be collected.In addiction, it is necessary to include more attributes such as medical conditions and disase parameters in order to improve accuracy. 


